NTM CT Scoring Agreement

### 2025-07-31 Stacey Martiniano/ntm_agreement.qmd
feature scores

```{r}
#| include: false
library(tidyverse)
library(redcapAPI)
library(Hmisc)
library(knitr)
library(fastDummies)
library(lmerTest)
library(performance)
# REDCap
unlockREDCap(c(rcon = "NTM CT Scoring"),
  keyring = "API_KEYs",
  envir = globalenv(),
  url = "https://redcap.ucdenver.edu/api/"
)
# Home directory
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF",
  "Windows" = "C:/Users/Tim/OneDrive - The University of Colorado Denver/Vigers/CF",
  "Linux" = "/home/tim/OneDrive/Vigers/CF"
)
github_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Documents/GitHub",
  "Windows" = "C:/Users/Tim/Documents/GitHub",
  "Linux" = "/home/tim/Documents/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
```

```{r data cleaning}
#| message: false
#| warning: false
# Import data from REDCap
data <- exportRecordsTyped(rcon, labels = F)
data <- data.frame(lapply(data, as.character))
data$scan_id <- as.numeric(data$scan_id)
# Make scorer IDs by initial
data$scorer <- paste0(
  toupper(substr(data$reader_first_name, 1, 1)),
  toupper(substr(data$reader_last_name, 1, 1))
)
data$scorer[data$scorer == "GG"] <- "GS"
# Deidentify readers
old <- c("AF", "DL", "JW")
new <- c("A", "B", "C")
data$scorer[data$scorer %in% old] <- new[match(data$scorer, old, nomatch = 0)]
# Revert duplicate IDs back to original to match with gold standard
training <- read.csv("./Stacey Martiniano/NTM CT Scoring/Data_Clean/training.key.csv")
dups <- read.csv("./Stacey Martiniano/NTM CT Scoring/Data_Clean/duplicated.key.csv")
training$Training_id <- as.numeric(sub("T00", "", training$Training_id))
# Change duplicates to OG IDs
old <- dups$duplicated_ssid..dates.shifted..17.days.
new <- dups$original_ssid
data$scan_id[data$scan_id %in% old] <- new[na.omit(match(data$scan_id, old))]
# Change training IDs to OG IDs
old <- training$Training_id
new <- training$Cffid.Scoring.Set
data$scan_id[data$scan_id %in% old] <- new[na.omit(match(data$scan_id, old))]
# Convert to long
data <- data %>%
  select(
    scan_id, scorer, scoresheet_timestamp,
    bronchiectasis_rul:atelectasis___6
  ) %>%
  arrange(scan_id, scorer, scoresheet_timestamp) %>%
  group_by(scan_id, scorer) %>%
  mutate(scorer_repeat = row_number()) %>%
  select(-scoresheet_timestamp) %>%
  pivot_longer(bronchiectasis_rul:atelectasis___6,
    names_to = c("Feature", "Location"),
    names_pattern = "(.*)_(.*)$"
  ) %>%
  mutate(
    Feature = gsub("__", "", Feature),
    Feature = str_to_title(gsub("_", " ", Feature))
  )
data$Location <- factor(data$Location,
  levels = c(
    "1", "2", "3", "4", "5", "6", "ling", "lll",
    "lul", "rll", "rml", "rul"
  ),
  labels = c(
    "RUL", "RML", "RLL", "LUL", "Ling", "LLL",
    "Ling", "LLL", "LUL", "RLL", "RML", "RUL"
  )
)
# Add tree in bud OR nodule variable
data <- data %>%
  pivot_wider(names_from = Feature) %>%
  mutate(
    `Tree In Bud OR Nodule` =
      factor(Nodule == "Checked" | `Tree In Bud` == "Checked",
        levels = c(F, T), labels = c("Unchecked", "Checked")
      )
  ) %>%
  pivot_longer(Bronchiectasis:last_col(), names_to = "Feature")
# Format values
data$value <- factor(data$value,
  levels = c(
    "Absent", "Checked", "Cylindric", "Cystic", "Unchecked", "Varicose"
  ),
  labels = c("Absent", "Present", "Present", "Present", "Absent", "Present"),
  ordered = T
)
# Separate out gold standards and duplicates
data_dups <- data %>% filter(scan_id %in% dups$original_ssid)
data_gs <- data %>% filter(scan_id %in% training$Cffid.Scoring.Set)
data <- data %>%
  filter(!scan_id %in% c(dups$original_ssid, training$Cffid.Scoring.Set))
# Second readings only for comparison to GS
data_gs <- data_gs %>%
  group_by(scan_id, scorer, Feature, Location) %>%
  slice_max(scorer_repeat)
```

@@

```{r}
# Import disease state
outcomes <- read.csv("./Stacey Martiniano/NTM CT Scoring/Data_Clean/disease.key.csv", na.strings = "")
outcomes$NTM.PD..Y.N. <- factor(outcomes$NTM.PD..Y.N.,
  levels = c("No", "Yes"), ordered = T
)
# For the duplicated scans, just use the first reading and combine with the
# singletons
scores <- rbind(data, data_dups[data_dups$scorer_repeat == 1, ])
# Add gold standards
scores <- rbind(scores, data_gs[data_gs$scorer == "GS", ])
# Add outcome info
scores <- left_join(scores,
  outcomes %>% select(cff_registry_id, NTM.PD..Y.N.),
  by = join_by(scan_id == cff_registry_id)
) %>%
  rename(`NTM-PD` = NTM.PD..Y.N.)
# Calculate the scores for each feature.
# When calculating the scores, take the majority vote by scan, feature, and
# location, since that's how they scored it.
# For example: If one scorer detected atelectasis in the RUL, one in the RML,
# and one in the RLL, but the majority at each location was "absent," then that
# scan would have a 0 for atelectasis. In other words, score the locations by
# majority first, then add the locations together for the feature score.
feature_scores <- scores %>% group_by(scan_id, Feature, `NTM-PD`)
```

# Mixed Model Approach

```{r data cleaning 2}
# Import data
data <- exportRecordsTyped(rcon, labels = F)
data <- data.frame(lapply(data, as.character))
data$scan_id <- as.numeric(data$scan_id)
# Make scorer IDs by initial
data$scorer <- paste0(
  toupper(substr(data$reader_first_name, 1, 1)),
  toupper(substr(data$reader_last_name, 1, 1))
)
data$scorer[data$scorer == "GG"] <- "GS"
# Deidentify readers
old <- c("AF", "DL", "JW")
new <- c("A", "B", "C")
data$scorer[data$scorer %in% old] <- new[match(data$scorer, old, nomatch = 0)]
# Revert duplicate IDs back to original to match with gold standard
training <- read.csv("./Stacey Martiniano/NTM CT Scoring/Data_Clean/training.key.csv")
dups <- read.csv("./Stacey Martiniano/NTM CT Scoring/Data_Clean/duplicated.key.csv")
training$Training_id <- as.numeric(sub("T00", "", training$Training_id))
# Change duplicates to OG IDs
old <- dups$duplicated_ssid..dates.shifted..17.days.
new <- dups$original_ssid
data$scan_id[data$scan_id %in% old] <- new[na.omit(match(data$scan_id, old))]
# Change training IDs to OG IDs
old <- training$Training_id
new <- training$Cffid.Scoring.Set
data$scan_id[data$scan_id %in% old] <- new[na.omit(match(data$scan_id, old))]
# Convert to long
data <- data %>%
  select(
    scan_id, scorer, scoresheet_timestamp,
    bronchiectasis_rul:atelectasis___6
  ) %>%
  arrange(scan_id, scorer, scoresheet_timestamp) %>%
  group_by(scan_id, scorer) %>%
  mutate(scorer_repeat = row_number()) %>%
  select(-scoresheet_timestamp) %>%
  pivot_longer(bronchiectasis_rul:atelectasis___6,
    names_to = c("feature", "location"),
    names_pattern = "(.*)_(.*)$"
  ) %>%
  mutate(
    feature = gsub("__", "", feature),
    feature = str_to_title(gsub("_", " ", feature))
  )
data$location <- factor(data$location,
  levels = c(
    "1", "2", "3", "4", "5", "6", "ling", "lll",
    "lul", "rll", "rml", "rul"
  ),
  labels = c(
    "RUL", "RML", "RLL", "LUL", "Ling", "LLL",
    "Ling", "LLL", "LUL", "RLL", "RML", "RUL"
  )
)
# Format values
data$value <- factor(data$value,
  levels = c(
    "Absent", "Checked", "Cylindric", "Cystic", "Unchecked", "Varicose"
  ),
  labels = c("Absent", "Present", "Present", "Present", "Absent", "Present")
)
data$value <- as.numeric(data$value) - 1
# Separate out gold standards and duplicates
data_dups <- data %>% filter(scan_id %in% dups$original_ssid)
data_gs <- data %>% filter(scan_id %in% training$Cffid.Scoring.Set)
data <- data %>%
  filter(!scan_id %in% c(dups$original_ssid, training$Cffid.Scoring.Set))
# Second readings only for comparison to GS
data_gs <- data_gs %>%
  group_by(scan_id, scorer, feature, location) %>%
  slice_max(scorer_repeat)
# Dummy variables for scorer, organize columns
data <- data %>%
  dummy_cols(select_columns = "scorer") %>%
  select(scan_id, feature, location, scorer, scorer_A, scorer_B, scorer_C, value)
```

## Data setup

```{r}
DT::datatable(data)
```

## Model

- Readers $r=1,2,3$
- Scans $s=1,2,...,S$
- Features $f=1,2,...,F$
- Lobes $l=1,2,...,6$
- Outcome $y_{rsfl}$ is binary
- Model notation: $$y_{rsf=1}=\alpha+\beta X_l + b_1+b_2+b_3+\epsilon_{rs}$$

### Bronchiectasis

```{r}
#| code-fold: show
df <- data %>% filter(feature == "Bronchiectasis")
model <- glmer(value ~ location + (1 | scorer),
  family = "binomial", data = df
)
summary(model)
icc(model)
```

### Mucoid Impaction

```{r}
#| code-fold: show
df <- data %>% filter(feature == "Mucoid Impaction")
model <- glmer(value ~ location + (1 | scorer),
  family = "binomial", data = df
)
summary(model)
icc(model)
```

### Cyst

```{r}
#| code-fold: show
df <- data %>% filter(feature == "Cyst")
model <- glmer(value ~ location + (1 | scorer),
  family = "binomial", data = df
)
summary(model)
icc(model)
```

### Cavity

```{r}
#| code-fold: show
df <- data %>% filter(feature == "Cavity")
model <- glmer(value ~ location + (1 | scorer),
  family = "binomial", data = df
)
summary(model)
icc(model)
```

### Consolidation

```{r}
#| code-fold: show
df <- data %>% filter(feature == "Consolidation")
model <- glmer(value ~ location + (1 | scorer),
  family = "binomial", data = df
)
summary(model)
icc(model)
```

### Ground Glass

```{r}
#| code-fold: show
df <- data %>% filter(feature == "Ground Glass")
model <- glmer(value ~ location + (1 | scorer),
  family = "binomial", data = df
)
summary(model)
icc(model)
```

### Nodule

```{r}
#| code-fold: show
df <- data %>% filter(feature == "Nodule")
model <- glmer(value ~ location + (1 | scorer),
  family = "binomial", data = df
)
summary(model)
icc(model)
```

### Tree In Bud

```{r}
#| code-fold: show
df <- data %>% filter(feature == "Tree In Bud")
model <- glmer(value ~ location + (1 | scorer),
  family = "binomial", data = df
)
summary(model)
icc(model)
```

### Atelectasis

```{r}
#| code-fold: show
df <- data %>% filter(feature == "Atelectasis")
model <- glmer(value ~ location + (1 | scorer),
  family = "binomial", data = df
)
summary(model)
icc(model)
```

>>>>

```{r}
#| include: false
library(tidyverse)
library(redcapAPI)
library(Hmisc)
library(knitr)
library(lmerTest)
library(performance)
# REDCap
unlockREDCap(c(rcon = "NTM CT Scoring"),
  keyring = "API_KEYs",
  envir = globalenv(),
  url = "https://redcap.ucdenver.edu/api/"
)
# Home directory
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF",
  "Windows" = "C:/Users/Tim/OneDrive - The University of Colorado Denver/Vigers/CF",
  "Linux" = "/home/tim/OneDrive/Vigers/CF"
)
github_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Documents/GitHub",
  "Windows" = "C:/Users/Tim/Documents/GitHub",
  "Linux" = "/home/tim/Documents/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
```

```{r data cleaning}
#| message: false
#| warning: false
# Import data from REDCap
data <- exportRecordsTyped(rcon)
data <- data.frame(lapply(data, as.character))
data$scan_id <- as.numeric(data$scan_id)
# Make scorer IDs by initial
data$scorer <- paste0(
  toupper(substr(data$reader_first_name, 1, 1)),
  toupper(substr(data$reader_last_name, 1, 1))
)
data$scorer[data$scorer == "GG"] <- "GS"
# Deidentify readers
old <- c("AF", "DL", "JW")
new <- c("A", "B", "C")
data$scorer[data$scorer %in% old] <- new[match(data$scorer, old, nomatch = 0)]
# Revert duplicate IDs back to original to match with gold standard
training <- read.csv("./Stacey Martiniano/NTM CT Scoring/Data_Clean/training.key.csv")
dups <- read.csv("./Stacey Martiniano/NTM CT Scoring/Data_Clean/duplicated.key.csv")
training$Training_id <- as.numeric(sub("T00", "", training$Training_id))
# Change duplicates to OG IDs
old <- dups$duplicated_ssid..dates.shifted..17.days.
new <- dups$original_ssid
data$scan_id[data$scan_id %in% old] <- new[na.omit(match(data$scan_id, old))]
# Change training IDs to OG IDs
old <- training$Training_id
new <- training$Cffid.Scoring.Set
data$scan_id[data$scan_id %in% old] <- new[na.omit(match(data$scan_id, old))]
# Convert to long
data <- data %>%
  select(
    scan_id, scorer, scoresheet_timestamp,
    bronchiectasis_rul:atelectasis___6
  ) %>%
  arrange(scan_id, scorer, scoresheet_timestamp) %>%
  group_by(scan_id, scorer) %>%
  mutate(scorer_repeat = row_number()) %>%
  select(-scoresheet_timestamp) %>%
  pivot_longer(bronchiectasis_rul:atelectasis___6,
    names_to = c("Feature", "Location"),
    names_pattern = "(.*)_(.*)$"
  ) %>%
  mutate(
    Feature = gsub("__", "", Feature),
    Feature = str_to_title(gsub("_", " ", Feature))
  )
data$Location <- factor(data$Location,
  levels = c(
    "1", "2", "3", "4", "5", "6", "ling", "lll",
    "lul", "rll", "rml", "rul"
  ),
  labels = c(
    "RUL", "RML", "RLL", "LUL", "Ling", "LLL",
    "Ling", "LLL", "LUL", "RLL", "RML", "RUL"
  )
)
# Add tree in bud OR nodule variable
data <- data %>%
  pivot_wider(names_from = Feature) %>%
  mutate(
    `Tree In Bud OR Nodule` =
      factor(Nodule == "Checked" | `Tree In Bud` == "Checked",
        levels = c(F, T), labels = c("Unchecked", "Checked")
      )
  ) %>%
  pivot_longer(Bronchiectasis:last_col(), names_to = "Feature")
# Format values
data$value <- factor(data$value,
  levels = c(
    "Absent", "Checked", "Cylindric", "Cystic", "Unchecked", "Varicose"
  ),
  labels = c("Absent", "Present", "Present", "Present", "Absent", "Present"),
  ordered = T
)
# Separate out gold standards and duplicates
data_dups <- data %>% filter(scan_id %in% dups$original_ssid)
data_gs <- data %>% filter(scan_id %in% training$Cffid.Scoring.Set)
data <- data %>%
  filter(!scan_id %in% c(dups$original_ssid, training$Cffid.Scoring.Set))
# Second readings only for comparison to GS
data_gs <- data_gs %>%
  group_by(scan_id, scorer, Feature, Location) %>%
  slice_max(scorer_repeat)
```

@@

## Total disease feature score (orange)

```{r}
# Import disease state
outcomes <-
  read.csv("./Stacey Martiniano/NTM CT Scoring/Data_Clean/disease.key.csv",
    na.strings = ""
  )
outcomes$NTM.PD..Y.N. <- factor(outcomes$NTM.PD..Y.N.,
  levels = c("No", "Yes"), ordered = T
)
outcomes <- outcomes %>%
  select(cff_registry_id, NTM.PD..Y.N.) %>%
  rename(`NTM-PD` = NTM.PD..Y.N.)
# For the duplicated scans, just use the first reading and combine with the
# singletons
scores <- rbind(data, data_dups[data_dups$scorer_repeat == 1, ])
# Add gold standards
scores <- rbind(scores, data_gs[data_gs$scorer == "GS", ])
# Calculate the scores for each feature.
# When calculating the scores, take the majority vote by scan, feature, and
# location, since that's how they scored it.
# For example: If one scorer detected atelectasis in the RUL, one in the RML,
# and one in the RLL, but the majority at each location was "absent," then that
# scan would have a 0 for atelectasis. In other words, score the locations by
# majority first, then add the locations together for the feature score.
feature_scores <- scores %>%
  group_by(scan_id, Feature, Location) %>%
  summarise(Majority = factor(sum(value == "Present"),
    levels = 0:3, labels = c("Absent", "Absent", "Present", "Present")
  ), .groups = "drop_last") %>%
  summarise(Score = sum(Majority == "Present"), .groups = "drop")
# Add outcome data
feature_scores <- left_join(feature_scores, outcomes,
  by = join_by(scan_id == cff_registry_id)
)
# Summary table
feature_scores %>%
  pivot_wider(names_from = Feature, values_from = Score) %>%
  select(-scan_id) %>%
  tbl_summary(by = `NTM-PD`, missing_text = "Missing")
feature_scores %>%
  pivot_wider(names_from = Feature, values_from = Score) %>%
  select(-scan_id) %>%
  tbl_summary(
    by = `NTM-PD`, missing_text = "Missing",
    type = list(
      c(
        "Atelectasis", "Bronchiectasis", , "Cavity", "Consolidation", "Cyst",
        "Ground Glass", "Mucoid Impaction", "Nodule", "Tree In Bud",
        "Tree In Bud OR Nodule"
      ) ~ "continuous2"
    ),
    statistic = all_continuous() ~ c("{median} ({p25}, {p75})", "{min}, {max}")
  )
```
